Determinism, Complexity, and Predictability in Computer Performance

### 2014-07-14 response.tex

\emph{3) By the way, the authors call ``fully complex'' the white
  noise process, i.e. they equate predictability and
  complexity. However, they do not even mention another approach to
  the notion of complexity which ascribes low complexity to the white
  noise [the works e.g. by C.R. Shalizi and J. Crutchfield carried out
    in the Santa Fe Institute from where one of the authors is]. It
  would be appropriate to mention that ambiguity of the notion of
  complexity if the latter plays an important role in many
  formulations of the authors.}

We have added a short discussion about the differences in notions and
definitions of complexity, and clarified the working definition that
we use in this paper.

\emph{4) The second finding reads ``The way information is generated
  and processed internally by a system plays a crucial role in the
  success of different forecasting schema - and in the choice of which
  one is appropriate for a given time series''. This statement also
  looks trivial. Indeed, an optimal predictor for a linear stochastic
  system and low-dimensional nonlinear deterministic system are quite
  different (an AR model versus a local or global nonlinear model) due
  to different properties of the original systems. Thus, there is
  nothing to be proven here. Probably, the authors imply their
  particular benchmark signals (reflecting a computer behavior) but
  they do not claim that.  Thus, both primary findings seem to provide
  nothing new.}

The original wording of this finding did make it sound trivial.  Our
intent, as we have modified the text to reflect, is that the
appropriateness of a predictor can be discovered \emph{empirically},
via comparison with the WPE.  Previous explorations have used clean
data and/or systems where the generating process is known.  Our work
makes neither of these assumptions.  The paragraph that we have added
to the introduction (immediately following these two claims) should
hopefully make this more clear.  \alert{Ryan said: ``i'm not entirely
  sure how to respond to this one correctly.''  Liz asks: do you mean
  in the rebuttal or in the text?}

@@

Ryan had written: ``the specific relationship between WPE and MASE
values is dependent on word length, though for any reasonable choice
of $\ell$, we believe a similar relationship to hold.'' Liz is not
sure she understands what ``similar'' means or why that's justified.}

Regarding generality: the goal of this paper was to explore model-free
heuristics for empirical time-series data, where one does not know
anything about the underlying system.  In situations like this,
equivalence classes are difficult to define.  We addressed
this---insofar as is possible in an experimental context---by choosing
a range of time-series data sets whose behavior spans a wide range of
dynamical behaviors.  Recapitulate that henon, lorenz, SFI A all fit
the curve determined by the computer perf data; cite as justification
for generality and point to future work.

\emph{II) PE is invariant under an invertible nonlinear change of
  variables. The prediction error measured as the authors suggest is
  not invariant. Thus, such a simple change of variables can lead, at
  least, to a change in regression coefficient between WPE and
  prediction error. Thus, the authors should also discuss to which
  variable for that class of systems their result apply. I think that
  the fact that the authors observe a certain regression line at all
  seems to be due to small number of different signals they consider
  (only 7 or 8 different data points, such a small number of points
  may lie approximately near a straight line by chance).}

Both WPE and MASE are sensitive to changes of variables; our initial
testing, however, indicates they are so \emph{in complementary ways},
so the shape of the WPE vs. MASE curve is [[largely invariant with
    respect to changes in what]] \alert{Ryan wrote: ``This leads a
  change in variables to...''  What should go in here?  What variables
  is he talking about?  Did we actually do that test (varying them)?
  What should we say about that word ``largely''?}

The fit of the regression line was not a function of the small number
of points; when we included all 105 WPE-MASE pairs, rather than just a
representative from each program, the slope of the fit changed by 2\%
and the y intercept by 1\%.

We have revised the Figure, and the associated text, to include all
105 points.  We have also moved from a linear fit to a log fit in this
curve to better match the theoretical relationship between MASE and
WPE, as mentioned above.  Finally, we now show a ``fuzzy'' version of
the fitted curve in an effort to emphasize that the fundamental claim
of the paper is not related to the crispness of the fit.

### 2014-07-14 concl.tex
second round of revisions.  awaiting today's conversation before can finish.

Forecast strategies that are designed to capture predictive structure
are ineffective when signal complexity outweighs information
redundancy.  This poses a number of serious challenges in practice.
Without knowing anything about the generating process, it is difficult
to determine how much predictive structure is present in a given time
series.  And even if predictive structure exists, a given forecast
method may not work, simply because it cannot exploit the structure
that is present (e.g., a linear model of a nonlinear process).  If a
forecast model is not producing good results, a practitioner needs to
know why: is the reason that the time series contains no predictive
structure---i.e., that no model will work---or is the model that s/he
is using simply not good enough?

In this paper, we have argued that redundancy~\cite{crutchfield2003}
is a useful definition of the inherent predictability of a time
series.  To operationalize that definition, we use an approximation of
the Kolmogorov-Sinai entropy~\cite{lind95}, estimated using a weighted
version of the permutation entropy of~\cite{bandt2002per}.  This WPE
technique---an ordinal calculation of forward information transfer in
a time series---is ideal for our purposes because it works with
real-valued data and is known to converge to the true entropy value.
Using a variety of forecast models and more than a hundred time-series
data sets from a computer-performance experiment, we have shown that
prediction accuracy is indeed correlated with weighted permutation
entropy: the higher the WPE, in general, the higher the prediction
error.  There are some exceptions: \svdone is easier to predict than
its WPE value suggests using any of the three forecast strategies
studied here.  Conversely, forecast errors for the \col and \svdfive
signals are higher than the corresponding WPEs suggest---except for
the nonlinear LMA predictor.  On the whole, though, signals with
higher weighted permutation entropy are harder to predict.

@@

An important practical corollary to this empirical correlation of
predictability and WPE is a practical strategy for assessing
appropriateness of forecast methods.  If the forecast produced by a
particular method is poor but the time series contains a significant
amount of predictive structure, one can reasonably conclude that that
method is inadequate to the task and that one should seek another
method.  The nonlinear LMA method, for instance, performs better in
most cases because it is more general.  (This is particularly apparent
in the \col and \svdfive examples.)
% , where the other two methods do not perform well.)  
The \naive ~method, which simply predicts the mean, can work very well
on noisy signals because it effects a filtering operation.  The simple
random-walk strategy outperforms LMA, ARIMA, and the \naive ~method on
the \gcc signal, which is extremely complex---i.e., extremely low
redundancy.
%  \naive ~wins on \svdone with the exception to the 5 outliers.
The dashed line in Figures~\ref{fig:wpe_vs_mase_best}
and~\ref{fig:wpe_vs_mase_all} operationalizes the discussion in this
paragraph.  It is a preliminary, but potentially useful, empirical
guideline for knowing when a model is not well-matched to the task at
hand: a MASE score that is more than $3.58 \cdot \mathrm{WPE} - 1.91$,
where $\mathrm{WPE}$ is that of the corresponding signal, is an
indication that that time series has more predictive structure than
that forecast model can capture and exploit.

>>>>

Forecast strategies that are designed to capture predictive structure
are ineffective when signal complexity outweighs information
redundancy.  This poses a number of serious challenges in practice.
Without knowing anything about the generating process, it is difficult
to determine how much predictive structure is present in a noisy,
real-world time series.  And even if predictive structure exists, a
given forecast method may not work, simply because it cannot exploit
the structure that is present (e.g., a linear model of a nonlinear
process).  If a forecast model is not producing good results, a
practitioner needs to know why: is the reason that the data contain no
predictive structure---i.e., that no model will work---or is the model
that s/he is using simply not good enough?

In this paper, we have argued that redundancy~\cite{crutchfield2003}
is a useful definition of the inherent predictability of an empirical
time series.  To operationalize that definition, we use an
approximation of the Kolmogorov-Sinai entropy~\cite{lind95}, estimated
using a weighted version of the permutation entropy
of~\cite{bandt2002per}.  This WPE technique---an ordinal calculation
of forward information transfer in a time series---is ideal for our
purposes because it works with real-valued data and is known to
converge to the true entropy value. Using a variety of forecast models
and more than a hundred time-series data sets from a
computer-performance experiment, we have shown that prediction
accuracy is indeed correlated with weighted permutation entropy: the
higher the WPE, in general, the higher the prediction error.  The
relationship is roughly logarithmic, which makes sense because any
signal with zero entropy should be perfectly predictable, and so
should ideally have a MASE value of zero.  Supporting our hypothesis,
forecast errors for the \col and \svdfive signals are higher than the
corresponding WPEs suggest---except for the nonlinear LMA predictor.
There are some exceptions: \svdone is easier to predict than its WPE
value suggests using any of the three forecast strategies studied
here.  On the whole, though, signals with higher weighted permutation
entropy are harder to predict.

@@

An important practical corollary to this empirical correlation of
predictability and WPE is a practical strategy for assessing
appropriateness of forecast methods.  If the forecast produced by a
particular method is poor but the time series contains a significant
amount of predictive structure, one can reasonably conclude that that
method is inadequate to the task and that one should seek another
method.  The nonlinear LMA method, for instance, performs better in
most cases because it is more general.  (This is particularly apparent
in the \col and \svdfive examples.)
% , where the other two methods do not perform well.)
The \naive ~method, which simply predicts the mean, can work very well
on noisy signals because it effects a filtering operation.  The simple
random-walk strategy outperforms LMA, ARIMA, and the \naive ~method on
the \gcc signal, which is extremely complex---i.e., extremely low
redundancy.
%  \naive ~wins on \svdone with the exception to the 5 outliers.
The curve in Figures~\ref{fig:wpe_vs_mase_best}
and~\ref{fig:wpe_vs_mase_all} operationalizes the discussion in this
paragraph.  It is a preliminary, but potentially useful, empirical
guideline for knowing when a model is not well-matched to the task at
hand: a MASE score that is significantly higher than [[$3.58 \cdot
    \mathrm{WPE} - 1.91$]], where $\mathrm{WPE}$ is that of the
corresponding signal, is an indication that that time series has more
predictive structure than that forecast model can capture and exploit.

[[This paragraph duplicates material in the last paragraph of this
    section.  Some of it needs to be here rather than there, judging
    by the ref's reactions.  But the later paragraph has a nice
    concluding tone, so need to figure out what goes where.]]

This curve was determined empirically using a finite set of forecast
methods and data sets---the latter drawn from a single experimental
system.  Computer dynamics spans the full range of behavior reported
in the nonlinear dynamics literature.  Standard nonlinear time-series
analysis of data from these systems has confirmed the existence of
fixed points, periodic orbits of different periods, chaotic
trajectories with various Lyapunov exponents, etc. in their dynamics.
While it is certainly possible that the nonlinear dynamics of computer
performance is subtly different from the nonlinear dynamics of other
systems---in a manner that cannot be detected by traditional nonlinear
time-series analysis---we believe that our results will hold for time
series drawn from other physical systems.  We are certainly aware that
this claim may not be true, however, and we are in the process of a
follow-up study exploring a broader range of applications.
Preliminary results of this study show that MASE vs. WPE points for
data from the SFI prediction competition \cite{sfi} fall in the shaded
region on Figure [[what]]\footnote{Synthetic data from paradigmatic
  systems like the Henon map \cite{henon} and the Lorenz system
  \cite{lorenz63} also fit this pattern.}.

### 2014-07-14 results.tex
more updates
second round of revisions.  awaiting today's conversation before can finish.

\item That the weighted permutation entropy (WPE) of a noisy
  real-valued time series is correlated with prediction
  accuracy---i.e., that the predictable structure in a time-series
  data set can be quantified by its WPE.

@@

\item That the way information is generated and processed by a system
  is correlated with the effectiveness of a given predictor on
  time-series data from that system

@@

The experiments below involve three different prediction methods
applied to time-series data from eight different systems: {\tt
  col\_major}, {\tt 403.gcc}, and the six different regimes of the
{\tt dgesdd} signal in Figure~\ref{fig:svd-ts-colored}.  The objective of these
experiments was to explore how prediction accuracy is related to WPE,
and how that relationship depends on the generating process and the
prediction method.  Working from the first 90\% of each signal, we
generated a prediction of the last 10\% using the \naive, ARIMA, and
LMA prediction methods, as described in Section~\ref{sec:accuracy},
then calculated the MASE value of those predictions.  We also
calculated the WPE of each time series using a wordlength of six, as
described in Section~\ref{sec:meaComplex}.  In order to assess the
run-to-run variability of these results, we repeated all of these
calculations on 15 separate trials: i.e., 15 different runs of each
program.

Figure~\ref{fig:wpe_vs_mase_best} shows the \emph{best} of these 45
predictions for each system: that is, the lowest error over all 15
trials and all three methods for each of the eight programs.  The WPE
is plotted against the corresponding MASE value here in order to bring
out the correlation between these two quantities.
\begin{figure}[htbp]
  \centering
  \includegraphics[width=\columnwidth]{figs/prediction_vs_entropy}
  \caption{Weighted permutation entropy vs. mean absolute scaled error
    (MASE) of the best prediction of each system.
% 
% (across all trials and all prediction methods) 
% 
The dashed line is a least-squares linear fit of all the points except
{\tt dgesdd$_1$}, which we have excluded for reasons explained in the
text.}
  \label{fig:wpe_vs_mase_best}
\end{figure}
As the plot shows, the best-case prediction error for each of these
eight systems is roughly proportional to the weighted permutation
entropy, which is consistent with our first conjecture.  The dashed
line in the figure, a linear least-squares fit to these best-case
errors\footnote{with the exception of {\tt dgesdd$_1$}, for reasons
  described below} captures this rough proportionality.  This is not a
formal result.  The three forecast methods used here were chosen to
span the space of standard prediction strategies, but they do not
cover that space exhaustively.  Our goal here is an empirical
assessment of predictability and complexity, not formal results about
a ``best'' predictor for a given time series.  There may be other
methods that produce lower MASE values than those in
Figure~\ref{fig:wpe_vs_mase_best}, but the sparseness of the points in
the upper-left and lower-right quadrants of this plot strongly
suggests that the underlying predictability of a time series is
inversely correlated with its WPE.  The rest of this section describes
our results in more detail---including the measures taken to assure
meaningful comparisons across methods, trials, and programs---and
elaborates on the meaning of the dashed line in the figure.

Figure~\ref{fig:wpe_vs_mase_all} shows WPE vs. MASE plots for the full
set of experiments.
\begin{figure}
  \centering
%    \includegraphics[width=\columnwidth]{figs/ARIMA_prediction_vs_entropy}
%    \includegraphics[width=\columnwidth]{figs/ARIMA_prediction_vs_entropy}
%    \includegraphics[width=\columnwidth]{figs/LMA_prediction_vs_entropy}
  \includegraphics[width=\columnwidth]{figs/predictions_vs_entropy3}
\caption{WPE vs. MASE for all trials, methods, and systems---with the
  exception of \svdone, \svdthree, and \svdfive are omitted from the
  top plot for scale reasons.
% 
% ; their MASE scores are $2.676 \pm 4.328$, $31.39 \pm 0.282$, and
%   $20.87 \pm 0.192$ respectively. 
% 
Numerical values, including means and standard deviations of the
errors, can be found in Table~\ref{tab:error}.  The dashed line is the
same as in the previous figure.
}
    \label{fig:wpe_vs_mase_all}
\end{figure} 
There are 15 points in each cluster, one for each trial.  (The points
in Figure~\ref{fig:wpe_vs_mase_best} are the leftmost of the points
for the corresponding system in any of the three plots in
Figure~\ref{fig:wpe_vs_mase_all}.)  The WPE values do not vary very
much across trials.  For most traces, the variance in MASE scores is
low as well, resulting in small, tight clusters.  In some
cases---ARIMA predictions of {\tt col\_major}, for instance---the MASE
variance is larger, which spreads out the clusters horizontally.  The
distribution of the cluster \emph{positions} is quite different in the
three plots.  The mean MASE scores of predictions generated with the
nonlinear LMA method are generally closer to the dashed line; the
ARIMA method clusters are more widely spread, the \naive ~clusters
even more so.  This geometric validation of the second conjecture in
this paper is discussed in more detail in the following paragraphs.

@@

This disparity---WPE values that suggest a high rate of forward
information transfer in the signal, but predictions with comparatively
poor MASE scores---is obvious in the geometry of the top two images
in Figure~\ref{fig:wpe_vs_mase_all}, where the \col clusters are far
to the right of the dashed line.  \emph{This indicates that these
  methods are not leveraging the available information in the signal}.
The dynamics of \col may be complicated, but they are not
unstructured.  This signal is nonlinear and deterministic\cite{mytkowicz09}, and if one
uses a prediction technique that is based a nonlinear model
(LMA)---rather than a method that simply predicts the running mean
(\naive) or one that uses a linear model (ARIMA)---the MASE score is
much better: $0.050 \pm 0.001$.  This prediction is 20 times more
accurate than a random-walk forecast, which is more in line with the
level of predictive structure that the low WPE value suggests is
present in the signal.

The dashed line in Figures~\ref{fig:wpe_vs_mase_best}
and~\ref{fig:wpe_vs_mase_all} is a heuristic that captures the
argument at the end of the previous paragraph.  It can allow
practitioners to recognize when a prediction method is not well
matched to the task at hand: that is, when the time series has more
predictive structure than the method is able to use.  When the
predictions are well below that line (as in the top and middle images
in Figure~\ref{fig:wpe_vs_mase_all}), one should try a different
method.  Conversely, the position of the LMA cluster---well to the
left of the other two, near the dashed line---reflects this method's
ability to capture and exploit the structure that is present in this
signal.

Note that the \col points (blue {\color{blue}$+$} icons) are clustered
very tightly in the lower left quadrant of the LMA and \naive ~MASE
vs. WPE plots, but spread out horizontally in the ARIMA plot.  This is
because of the way that these models are built \cite{autoARIMA}.  If
a KPSS test of the time series in question indicates that it is
nonstationary, an integration term is added to the model.  This test
gives mixed results in the case of the \col process, flagging five of
the 15 trials as stationary and ten as nonstationary.  ARIMA models
without an integration term perform more poorly on this signal, which
increases the error, thereby spreading out the points\footnote{We
  tested this hypothesis by forcing the inclusion of an integration
  term in the five cases where a KPSS test indicated that such a term
  was not needed.  This action removed the spread, pushing all 15 of
  the \col ARIMA points in Figure~\ref{fig:wpe_vs_mase_all} into a
  tight cluster.}.

The WPE of \svdfive ($0.677 \pm 0.006$) is higher than that of {\tt
  col\_major}.  This indicates that the rate of forward information
transfer of the underlying process is lower, but that time-series data
from this system still contain a significant amount of structure that
can, in theory, be leveraged to predict the future course of the time
series.  As mentioned above, though, the effectiveness of any
prediction strategy depends on how well it leverages the available
information: methods that use linear models, for instance, may fail
when applied to nonlinear processes.
% 
% (This is the basis for the second conjecture above.)  
% 
This match between model and task is an issue in the \svdfive
experiments as well.  The MASE scores of the \naive ~and ARIMA
predictions for this system are $20.870 \pm 0.192$ and $2.370 \pm
0.051$, respectively: that is, 20.87 and 2.37 times worse than a
simple random walk forecast of the same signals\footnote{The \naive
  ~MASE score is large because of the bimodal nature of the
  distribution of the values of the signal, which makes guessing the
  mean a particularly bad strategy.  The same thing is true of the
  \svdthree signal.}.  As before, the positions of these points on a
WPE vs. MASE plot---significantly below and to the right of the
dashed line---should suggest to a practitioner that there is more
structure in this signal than the corresponding method is able to
leverage, and that a different method might do better.  Indeed, for
\svdfive, the LMA method produces a MASE score of $ 0.718\pm 0.048 $
and a cluster of results that is much closer to the dashed line on the
WPE-MASE plot.  Again, this validates our second conjecture: the LMA
method can capture and reproduce the way in which the \svdfive system
processes information, but the \naive ~and ARIMA prediction methods
cannot.

@@

The effects described in the previous paragraph are exacerbated by the
way MASE is calculated.  Recall that MASE scores are scaled
\emph{relative to a random-walk forecast}, which uses the current
value as the prediction.  This strategy works very badly on signals
with frequent, large, rapid transitions.  Consider a signal that
oscillates from one end of its range to the other at every step.  A
signal like this will have a low WPE, much like {\tt col\_major}.
However, a random-walk forecast of this signal will be 180 degrees out
of phase with the true continuation.  Since random-walk error appears
in the denominator of the MASE score, this effect can shift points
leftwards on a WPE vs. MASE plot, and that is exactly why the
\svdone clusters for ARIMA and LMA are above the dashed line in
Figure~\ref{fig:wpe_vs_mase_all}.  This time series, which is shown in
closeup in Figure~\ref{fig:svdone-ts},
\begin{figure}[htbp]
  \centering
    \includegraphics[width=\columnwidth]{figs/svdonets2}
\caption{A small portion of the \svdone time series}\label{fig:svdone-ts}
\end{figure} 
is not quite to the level of the worst-case signal described above,
but it still poses a serious challenge to random-walk prediction.  It
is dominated by a noisy regime (between $\approx$1.86 and
$\approx$1.88 on the vertical scale in Figure~\ref{fig:svdone-ts}),
punctuated by short excursions above 1.9.  In the former regime, which
makes up more than 80\% of the signal, there are frequent dips to 1.82
and occasional larger dips below 1.8.  These single-point dips are the
bane of random-walk forecasting.  In this particular case, roughly
40\% of the forecasted points are off by the width of the associated
dip, which skews the associated MASE scores.  Signals like this are
also problematic for the \naive ~prediction strategy, since the
outliers have significant influence on the mean.  This compounds the
effect of the skew in the scaling factor and exacerbates the spread in
the \svdone MASE values.  For all of these reasons, we view \svdone as
an outlier and exclude it from the fit calculation of the dashed line
in Figure~\ref{fig:wpe_vs_mase_best}.

>>>>

\item That the weighted permutation entropy (WPE) of a noisy
  real-valued time series from an unknown system is correlated with
  prediction accuracy---i.e., that the predictable structure in an
  empirical time-series data set can be quantified by its WPE.

@@

\item That the way information is generated and processed by an
  unknown system is correlated with the effectiveness of a given
  predictor on empirical time-series data from that system

@@

The experiments below involve three different prediction methods
applied to time-series data from eight different systems: {\tt
  col\_major}, {\tt 403.gcc}, and the six different regimes of the
{\tt dgesdd} signal in Figure~\ref{fig:svd-ts-colored}.  The objective
of these experiments was to explore how prediction accuracy is related
to WPE.
% 
% , and how that relationship depends on the generating process
% and the prediction method.  
% 
Working from the first 90\% of each signal, we generated a prediction
of the last 10\% using the \naive, ARIMA, and LMA prediction methods,
as described in Section~\ref{sec:accuracy}, then calculated the MASE
value of those predictions.  We also calculated the WPE of each time
series using a wordlength chosen via the procedure described in
Section~\ref{sec:meaComplex}.  In order to assess the run-to-run
variability of these results, we repeated all of these calculations on
15 separate trials: i.e., 15 different runs of each program.

% Figure~\ref{fig:wpe_vs_mase_best} shows the \emph{best} of these
% predictions for each trial of each system: that is, the lowest error
% over all three methods for each of the eight programs.  The WPE is
% plotted against the corresponding MASE value here in order to bring
% out the correlation between these two quantities.

Figure~\ref{fig:wpe_vs_mase_best} plots the WPE values versus the
corresponding MASE values [[for all of these trials.]]
\begin{figure}[htbp]
  \centering
  \includegraphics[width=\columnwidth]{figs/new_prediction_vs_entropy_autolog}
  \caption{Weighted permutation entropy vs. mean absolute scaled error
    (MASE) of [[all trials for each system]].
%
% (across all trials and all prediction methods)
%
The dashed curve is a least-squares log fit of all the points except
those from {\tt dgesdd$_1$}, which we have excluded for reasons
explained in the text.  [[Say what the colored region is and why we
    included it.]]}
  \label{fig:wpe_vs_mase_best}
\end{figure}
As the plot shows, the [[best-case]] prediction error for each of
these eight systems is roughly proportional to the log of the weighted
permutation entropy, which is consistent with our first conjecture.
The dashed line in the figure, a least-squares fit of $y = a \log(b x
+ c)$ to these errors\footnote{with the exception of {\tt dgesdd$_1$},
  for reasons described below} captures this rough
proportionality\footnote{The logarithmic fit is appropriate here
  because any signal with zero entropy should be perfectly
  predictable, and so should ideally have a MASE value of zero.}.
This is not a formal result.  The forecast methods and data sets used
here were chosen to span the space of standard prediction strategies
and the range of dynamical behaviors, but they do not cover those
spaces exhaustively.  Our goal here is an\emph{empirical} assessment
of the relationship between predictability and complexity, not formal
results about a ``best'' predictor for a given time series.  There may
be other methods that produce lower MASE values than those in
Figure~\ref{fig:wpe_vs_mase_best}, but the sparseness of the points
above and below the curve in this plot strongly suggests a pattern of
(inverse) correlation between the underlying predictability of a time
series and its WPE.  The rest of this section describes our results in
more detail---including the measures taken to assure meaningful
comparisons across methods, trials, and programs---and elaborates on
the meaning of the dashed curve in the figure.

Figure~\ref{fig:wpe_vs_mase_all} shows WPE vs. MASE plots for the full
set of experiments.
\begin{figure}
  \centering
%    \includegraphics[width=\columnwidth]{figs/ARIMA_prediction_vs_entropy}
%    \includegraphics[width=\columnwidth]{figs/ARIMA_prediction_vs_entropy}
%    \includegraphics[width=\columnwidth]{figs/LMA_prediction_vs_entropy}
  \includegraphics[width=\columnwidth]{figs/new_predictions_vs_entropy3}
\caption{WPE vs. MASE for all trials, methods, and systems---with the
  exception of \svdone, \svdthree, and \svdfive are omitted from the
  top plot for scale reasons.
%
% ; their MASE scores are $2.676 \pm 4.328$, $31.39 \pm 0.282$, and
%   $20.87 \pm 0.192$ respectively.
%
Numerical values, including means and standard deviations of the
errors, can be found in Table~\ref{tab:error}.  The dashed curve is the
same as in the previous figure.
}
    \label{fig:wpe_vs_mase_all}
\end{figure}
There are 15 points in each cluster, one for each trial.  [[(The
    points in Figure~\ref{fig:wpe_vs_mase_best} are the leftmost of
    the points for the corresponding system in any of the three plots
    in Figure~\ref{fig:wpe_vs_mase_all}.)]]  The WPE values do not
vary very much across trials.  For most traces, the variance in MASE
scores is low as well, resulting in small, tight clusters.  In some
cases---ARIMA predictions of {\tt col\_major}, for instance---the MASE
variance is larger, which spreads out the clusters horizontally.  The
distribution of the cluster \emph{positions} is quite different in the
three plots.  The mean MASE scores of predictions generated with the
nonlinear LMA method are generally closer to the dashed curve; the
ARIMA method clusters are more widely spread, the \naive ~clusters
even more so.  This geometric validation of the second conjecture in
this paper is discussed in more detail in the following paragraphs.

@@

This disparity---WPE values that suggest a high rate of forward
information transfer in the signal, but predictions with comparatively
poor MASE scores---is obvious in the geometry of the top two images in
Figure~\ref{fig:wpe_vs_mase_all}, where the \col clusters are far to
the right of the dashed curve.  \emph{This indicates that these
  methods are not leveraging the available information in the signal}.
The dynamics of \col may be complicated, but they are not
unstructured.  This signal is nonlinear and
deterministic\cite{mytkowicz09}, and if one uses a prediction
technique that is based a nonlinear model (LMA)---rather than a method
that simply predicts the running mean (\naive) or one that uses a
linear model (ARIMA)---the MASE score is much better: $0.050 \pm
0.001$.  This prediction is 20 times more accurate than a random-walk
forecast, which is more in line with the level of predictive structure
that the low WPE value suggests is present in the signal.

The dashed curve in Figures~\ref{fig:wpe_vs_mase_best}
and~\ref{fig:wpe_vs_mase_all} is a heuristic that captures the
argument at the end of the previous paragraph.  It can allow
practitioners to recognize when a prediction method is not well
matched to the task at hand: that is, when the time series has more
predictive structure than the method is able to use.  When the
predictions are well below that curve (as in the top and middle images
in Figure~\ref{fig:wpe_vs_mase_all}), one should try a different
method.  Conversely, the position of the LMA cluster---well to the
left of the other two, near the dashed curve---reflects this method's
ability to capture and exploit the structure that is present in this
signal.  [[Again, this curve was determined from a finite set of
    methods and data.  But we worked hard to make these experiments
    representative and comprehensive.  Forecast methods from simple to
    sophisticated; time-series data from a system whose behavior spans
    the dynamical behavior space.  While we are cautiously opmistic
    about its generality, more exploration will be required to firm up
    that conclusion.  Say that Henon, Lorenz and SFI A fall within the
    colored region too, and that we're working on a broader study.]]

Note that the \col points (blue {\color{blue}$+$} icons) are clustered
very tightly in the lower left quadrant of the LMA and \naive ~MASE
vs. WPE plots, but spread out horizontally in the ARIMA plot.  This is
because of the way that these models are built \cite{autoARIMA}.  If a
KPSS test of the time series in question indicates that it is
nonstationary, an integration term is added to the model.  This test
gives mixed results in the case of the \col process, flagging five of
the 15 trials as stationary and ten as nonstationary.  ARIMA models
without an integration term perform more poorly on this signal, which
increases the error, thereby spreading out the points\footnote{We
  tested this hypothesis by forcing the inclusion of an integration
  term in the five cases where a KPSS test indicated that such a term
  was not needed.  This action removed the spread, pushing all 15 of
  the \col ARIMA points in Figure~\ref{fig:wpe_vs_mase_all} into a
  tight cluster.}.  \alert{This paragraph probably needs more, in
  response to the ref's gritch about fitting.}

The WPE of \svdfive ($0.677 \pm 0.006$) is higher than that of {\tt
  col\_major}.  This indicates that the rate of forward information
transfer of the underlying process is lower, but that time-series data
observed from this system still contain a significant amount of
structure that can, in theory, be leveraged to predict the future
course of the time series.  As mentioned above, though, the
effectiveness of any prediction strategy depends on how well it
leverages the information that is available in the data: methods that
use linear models, for instance, may fail when applied to nonlinear
processes.
%
% (This is the basis for the second conjecture above.)
%
This match between model and task is an issue in the \svdfive
experiments as well.  The MASE scores of the \naive ~and ARIMA
predictions for this system are $20.870 \pm 0.192$ and $2.370 \pm
0.051$, respectively: that is, 20.87 and 2.37 times worse than a
simple random walk forecast of the same signals\footnote{The \naive
  ~MASE score is large because of the bimodal nature of the
  distribution of the values of the signal, which makes guessing the
  mean a particularly bad strategy.  The same thing is true of the
  \svdthree signal.}.  As before, the positions of these points on a
WPE vs. MASE plot---significantly below and to the right of the shaded
region---should suggest to a practitioner that there is more structure
in this signal than the corresponding method is able to leverage, and
that a different method might do better.  Indeed, for \svdfive, the
LMA method produces a MASE score of $ 0.718\pm 0.048 $ and a cluster
of results that is much closer to the dashed curve on the WPE-MASE
plot.  Again, this validates our second conjecture: the LMA method can
capture and reproduce the way in which the \svdfive system processes
information, but the \naive ~and ARIMA prediction methods cannot.

@@

The effects described in the previous paragraph are also exacerbated
by the way MASE is calculated.  Recall that MASE scores are scaled
\emph{relative to a random-walk forecast}, which uses the current
value as the prediction.  This strategy works very badly on signals
with frequent, large, rapid transitions.  Consider a signal that
oscillates from one end of its range to the other at every step.  A
signal like this will have a low WPE, much like {\tt col\_major}.
However, a random-walk forecast of this signal will be 180 degrees out
of phase with the true continuation.  Since random-walk error appears
in the denominator of the MASE score, this effect can shift points
leftwards on a WPE vs. MASE plot, and that is exactly why the \svdone
clusters for ARIMA and LMA are above the dashed curve in
Figure~\ref{fig:wpe_vs_mase_all}.  This time series, which is shown in
closeup in Figure~\ref{fig:svdone-ts},
\begin{figure}[htbp]
  \centering
    \includegraphics[width=\columnwidth]{figs/svdonets2}
\caption{A small portion of the \svdone time series}\label{fig:svdone-ts}
\end{figure}
is not quite to the level of the worst-case signal described above,
but it still poses a serious challenge to random-walk prediction.  It
is dominated by a noisy regime (between $\approx$1.86 and
$\approx$1.88 on the vertical scale in Figure~\ref{fig:svdone-ts}),
punctuated by short excursions above 1.9.  In the former regime, which
makes up more than 80\% of the signal, there are frequent dips to 1.82
and occasional larger dips below 1.8.  These single-point dips are the
bane of random-walk forecasting.  In this particular case, roughly
40\% of the forecasted points are off by the width of the associated
dip, which skews the associated MASE scores.  Signals like this are
also problematic for the \naive ~prediction strategy, since the
outliers have significant influence on the mean.  This compounds the
effect of the skew in the scaling factor and exacerbates the spread in
the \svdone MASE values.  For all of these reasons, we view \svdone as
an outlier and exclude it from the fit calculation of the dashed curve
in Figure~\ref{fig:wpe_vs_mase_best}.

[[Short paragraph about how MASE has its drawbacks but is the best
    overall choice here because it is scale free and can compare
    across signals.  Point to the MASE paper for extensive evaluation
    and comparison wrt other metrics. We are in the process of
    repeating this study with other metrics, other forecast methods,
    and time-series data from other systems.  Segue to next section.]]
